---
title: "Estimation and Interpretation of Group Differences"
author: "<br><br><span style='font-size:25px;'><strong>Lesley Chapman Hannah, Ph.D., M.S.</strong></span><br>College of Graduate Studies<br>Northeast Ohio Medical University"

format:
  revealjs:
    theme: simple
    slide-number: true
    transition: fade
    chalkboard: true
    smaller: false
execute:
  echo: true
  warning: false
  message: false
editor: visual
---

```{r setup, echo=FALSE}
library(tidyverse)
library(broom)
library(scales)
library(epitools)

set.seed(2026)

n_treat <- 220
n_control <- 220

trial_df <- tibble(
  id = 1:(n_treat + n_control),
  treatment = c(rep("New therapy", n_treat), rep("Standard care", n_control)),
  response = c(
    rbinom(n_treat, 1, 0.68),
    rbinom(n_control, 1, 0.52)
  ),
  serious_ae = c(
    rbinom(n_treat, 1, 0.12),
    rbinom(n_control, 1, 0.20)
  ),
  egfr_change = c(
    rnorm(n_treat, mean = 4.8, sd = 9.5),
    rnorm(n_control, mean = 1.4, sd = 9.8)
  ),
  age = round(rnorm(n_treat + n_control, mean = 59, sd = 11)),
  bmi = round(rnorm(n_treat + n_control, mean = 30, sd = 5), 1)
)
```

### From Single-Group Estimation to Comparative Estimation

::: {style="font-size:0.63em"}
In the previous section, we discussed how estimation works for a single population quantity:

-   point estimates for one mean or one proportion\
-   sampling variability and standard error\
-   confidence intervals for a single parameter

In the current lecture we will focus on comparative estimation

> How different are two groups, and how should that difference be estimated and interpreted?

**Biomedical examples**: clinical trials, comparative effectiveness studies, and treatment evaluations are designed to estimate between-group effects
:::

------------------------------------------------------------------------

### Estimation for Comparing Groups

::: {style="font-size:0.66em"}
Central Goal: discuss statistical estimation for group comparisons

Discuss the following comparative estimands:

-   difference in sample means for a continuous outcome\
-   risk difference for a binary outcome\
-   risk ratio for a binary outcome\
-   odds and odds ratios for a binary outcome

Each comparative estimand follows the same general framework:

- estimate the outcome within each group\
- compare groups on an interpretable scale\
- use confidence intervals to judge both magnitude and precision
:::

-----------------------------------------------------------------

### Review: Statistical Estimation

::: {style="font-size:0.66em"}

- statistical estimation involves turning observed sample data into evidence about an unknown population quantity

- for estimand comparisons, estimation answers a question like this:

> based on the differences observed in the study sample, what is the best estimate of the corresponding treatment effect in the population?

Process inovlves the following:

1. define the population quantity of interest  
2. calculate a sample-based quantity that targets it  
3. interpret that quantity as an estimate of the underlying effect  


:::

---

### Estimating Differences Between Groups

::: {style="font-size:0.6em"}

- goal is to understand how two groups differ:

- treatment vs control  
- exposed vs unexposed  
- biomarker-positive vs biomarker-negative  

We are now estimating:

> a contrast between two population quantities

:::

---

### Defining the Estimand in Group Comparisons

::: {style="font-size:0.66em"}

Review:

- **estimand [what we want to learn]** : population quantity we aim to learn about

- **estimator [method used to estimate target]**: rule or formula used to approximate the estimand using sample data


In a two-group comparison, the estimand shows the relationship between two population quantities, such as:

- difference in population means  
- difference in population probabilities  
- ratio of population probabilities  
- ratio of population odds  

:::

---

### Defining How Groups Differ: Multiple Comparative Measures

::: {style="font-size:0.66em"}

In a two-group study, the data provide:

- an outcome in the treatment group  
- an outcome in the control group  

From these, we can estimate **how the groups differ**

But there is not just one way to define that difference.

Differences between groups [treatment effect] can be framed as:

- **absolute change** in outcome  
- **relative change** in outcome  
- comparison of **event vs non-event likelihood**  

Each of which defines a different population quantity [estimand] that the study is trying to estimate


:::
---

### Defining How Groups Differ: Multiple Comparative Measures

::: {style="font-size:0.66em"}

Multiple comparative measures can exist within a study:

- **mean difference** → estimates difference in population averages  
- **risk difference** → estimates absolute difference in probabilities  
- **risk ratio** → estimates relative difference in probabilities
- **odds ratio** → estimates difference on the odds scale  

What changes is:

> **the question we are asking about how the groups differ in the population**

:::
---

## The Structure of All Comparative Estimators

::: {style="font-size:0.66em"}

comparative measures follow a similar estimation structure:

1. estimate the outcome in each group  
2. combine those estimates into a comparison  
3. interpret that comparison as a population effect  



:::

---

### Mean Difference

$$
\widehat{\Delta} = \bar{x}_1 - \bar{x}_0
$$

::: {style="font-size:0.66em"}
 
Mean Difference:

- **mean difference** is the difference between the average outcome in two groups.

- $\bar{x}_1$: sample mean in the treatment group  
- $\bar{x}_0$: sample mean in the control group  

quantity is used to estimate the corresponding population mean difference:$\mu_1 - \mu_0$ - true difference in average outcome between the two groups in the full population


estimator of the population mean difference: $\mu_1 - \mu_0$


:::

---

### Risk Difference 

$$
\widehat{RD} = \hat{p}_1 - \hat{p}_0
$$

::: {style="font-size:0.6em"}

**risk difference** : sample-based estimate of the true absolute difference in event probability between groups in the population


For a binary outcome, each group has a population event probability:

- $p_1$: true probability of the event in the treatment group population  
- $p_0$: true probability of the event in the control group population  


study provides sample-based estimates because the population probabilities are unknown because the full population is not observed:

- $\hat{p}_1$: observed event proportion in the treatment group  
- $\hat{p}_0$: observed event proportion in the control group  


:::
---

### Risk Difference 

$$
\widehat{RD} = \hat{p}_1 - \hat{p}_0
$$

::: {style="font-size:0.6em"}

  

**Risk difference** compares these two probabilities and asks:

> how many probability units apart are the groups?

Therefore

$$
\widehat{RD} = \hat{p}_1 - \hat{p}_0
$$

is used to estimate:

> $RD = p_1 - p_0$

the true **difference in event probability** between treatment and control in the population

:::
---


### Risk Difference 

::: {style="font-size:0.6em"}

Example:

Consider a clinical trial comparing a new therapy to standard care, where the outcome is **treatment response** (yes/no)

Suppose the observed response probabilities in the sample are:

- treatment group: $\hat{p}_1 = 0.62$  
- control group: $\hat{p}_0 = 0.48$  



if the treatment response probability is 0.62 and the control response probability is 0.48, then the estimand risk difference is:

$$
\widehat{RD} = 0.62 - 0.48 = 0.14
$$

Interpretation: there is an estimated **14 percentage-point increase** in response probability under treatment

:::

---


### Risk Ratio 

$$
\widehat{RR} = \frac{\hat{p}_1}{\hat{p}_0}
$$

::: {style="font-size:0.66em"}

risk ratio : 

  - estimator of the population relative risk
  - quantifies how many times larger [or smaller] the outcome probability is in one group compared to another

risk ratio compares two estimated probabilities:

- **numerator**: $\hat{p}_1$  
  → outcome probability for the treatment group
  → e.x.: estimated probability of the outcome in the **treatment group**

- **denominator**: $\hat{p}_0$  
  → outcome probability for the control group
  → e.x.: estimated probability of the outcome in the **control group**

:::

---

### Risk Ratio 

$$
\widehat{RR} = \frac{\hat{p}_1}{\hat{p}_0}
$$

::: {style="font-size:0.66em"}

Question:

> how many times as likely is the outcome in the treatment group compared to the control group?

**Interpretation**

- $\widehat{RR} = 1$ → equal probabilities  
- $\widehat{RR} > 1$ → outcome more likely in treatment  
- $\widehat{RR} < 1$ → outcome less likely in treatment  
:::

---

### Odds and Odds Ratio 

$$
\widehat{OR} = \frac{\hat{p}_1/(1-\hat{p}_1)}{\hat{p}_0/(1-\hat{p}_0)}
$$

::: {style="font-size:0.6em"}


**odds ratio**

  - metric measures the association between two groups by          comparing their odds
  -  sample-based estimate of how the odds of an outcome differ    between groups in the population [expressed on an odds scale]

Note:

- probabilities describe frequency of events  
- odds describe event relative to non-event  

**Important distinction**

- **probability** ($p$): proportion of individuals who experience the event  
- **odds** ($p/(1-p)$): ratio of event probability to non-event probability  

:::
---

### Comparing Groups Using Different Measures

::: {style="font-size:0.66em"}

Each measure estimates a different type of population comparision between groups, and answer:

> how do outcomes differ between groups in the population?

Where:

- Mean difference → difference in averages  
- Risk difference → difference in probabilities  
- Risk ratio → relative probability  
- Odds ratio → relative odds  

:::

---

### Clinical Trial Context: Defining the Treatment Effect

::: {style="font-size:0.63em"}
Consider a randomized clinical study evaluating a **new therapy** compared to **standard care**

Study Goal: estimate how the treatment changes those outcomes in the target population

Key outcomes of interest include:

- **Clinical response** [binary]: did the patient respond to treatment?  
- **Serious adverse events** [binary]: did the patient experience a harmful event?  
- **Change in kidney function (eGFR)** [continuous]: how much did kidney function improve or decline?  


**Connection to estimation**

- outcomes are measured on different scales, the treatment effect must be defined and estimated differently:

- **mean difference** : estimates difference in average eGFR change  
- **risk difference** : estimates absolute difference in probability (response or adverse event)  
- **risk ratio** : estimates relative difference in probability  
- **odds ratio** : estimates difference on the odds scale  

:::

------------------------------------------------------------------------

### Dataset Overview

::: {style="font-size:0.6em"}

```{r, echo=FALSE}
trialdf <- trial_df |>
  slice_head(n = 6)

trialdf
```

-  each row represents one trial participant

-   `treatment` : identifies the treatment arm\
-   `response` : coded 1 for clinical response and 0 for no response\
-   `serious_ae` : coded 1 for a serious adverse event and 0 for none\
-   `egfr_change` : change in estimated glomerular filtration rate from baseline to follow-up

::: 

------------------------------------------------------------------------


### Difference in Sample Means

::: {style="font-size:0.6em"}
$$
\widehat{\Delta} = \bar{x}_1 - \bar{x}_0
$$

estimated mean difference between two groups

-   $\widehat{\Delta}$ : estimated population mean difference\
    - difference in eGFR change\ 
-   $\bar{x}_1$ : sample mean in group 1\
    - sample mean eGFR change in the **new therapy group**\
-   $\bar{x}_0$ : sample mean in group 0\
    - sample mean eGFR change in the **standard care group**

- positive value : average outcome is higher in group 1\
- negative value : average outcome is lower in group 1

estimated difference in **average change in eGFR** between treatment groups

Measure compares:

> average change in kidney function for patients receiving the new therapy  
> vs  
> average change in kidney function for patients receiving standard care  

::: 
-----------------------------------------------------------------

### Group Means for eGFR Change
::: {style="font-size:0.6em"}
```{r}
egfr_summary <- trial_df |>
  group_by(treatment) |>
  summarise(
    n = n(),
    mean_egfr_change = mean(egfr_change),
    sd_egfr_change = sd(egfr_change)
  )

egfr_summary
```

**Interpretation**

**Average outcome within a treatment group**

→ on average, patients receiving the new therapy improved by **5.65 eGFR units**
→ on average, patients receiving standard care improved by **1.01 eGFR units**


**Estimated treatment effect**

$$
\widehat{\Delta}=5.65−1.01=4.64
$$
Estimated difference in average kidney-function change between trial arms:

- patients receiving the new therapy improved **4.64 eGFR units** more than patients receiving standard care
- positive value means the new-therapy group had a larger average improvement than standard care
::: 
-----------------------------------------------------------------



### Plot of Mean eGFR Change by Group
::: {style="font-size:0.6em"}
```{r, echo=FALSE}
egfr_ci_df <- trial_df |>
  group_by(treatment) |>
  summarise(
    n = n(),
    mean_egfr_change = mean(egfr_change),
    sd_egfr_change = sd(egfr_change),
    se = sd_egfr_change / sqrt(n),
    lower = mean_egfr_change - qt(0.975, df = n - 1) * se,
    upper = mean_egfr_change + qt(0.975, df = n - 1) * se
  )

plot <- ggplot(egfr_ci_df, aes(x = treatment, y = mean_egfr_change)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.08) +
  labs(
    title = "Mean Change in eGFR With 95% Confidence Intervals",
    x = "",
    y = "Mean change in eGFR"
  ) +
  theme_bw()

plot
```
**Interpretation**

New therapy group

  - mean increase in eGFR is higher (around ~5–6 units)\
  - confidence interval is entirely above zero\
    - results suggests a consistent improvement in kidney function
    
Standard care group

  - mean change is smaller (around ~1 unit)\
  - confidence interval is wider and appears to include values near or slightly below zero\
    - results suggests greater uncertainty and likely little to no improvement

::: 
-----------------------------------------------------------------

### Estimated Risk Difference for Clinical Response
::: {style="font-size:0.6em"}


Goal: find the difference in response probabilities between the two treatment groups

$$
\hat{p}_{new}−\hat{p}_{standard}
$$

\hat{p}_{new}: proportion of patients who responded under the new therapy
\hat{p}_{standard}: proportion who responded under standard care

> How much does the probability of clinical response change between the two groups?

positive value → higher response probability under the new therapy
negative value → lower response probability under the new therapy
value near zero → little or no difference between groups

::: 
------------------------------------------------------------------------

### Estimated Risk Difference for Clinical Response
::: {style="font-size:0.6em"}
```{r, echo=FALSE}
response_summary <- trial_df |>
  group_by(treatment) |>
  summarise(
    n = n(),
    responses = sum(response),
    nonresponses = n() - responses,
    risk = mean(response)
  )

response_summary

risk_new <- response_summary$risk[response_summary$treatment == "New therapy"]
risk_std <- response_summary$risk[response_summary$treatment == "Standard care"]

risk_difference <- risk_new - risk_std
risk_difference
```

**Interpretation**

New therapy risk = 0.732
→ 73.2% of patients responded

Standard care risk = 0.495
→ 49.5% of patients responded

Out of 100 patients:

  ~ 73 would respond under the new therapy
  ~ 50 would respond under standard care
  
Risk Difference: $0.732−0.495=0.236$

probability of clinical response is 23.6 percentage points higher under the new therapy compared to standard care

::: 
------------------------------------------------------------------------





### Estimated Risk Ratio for Clinical Response
::: {style="font-size:0.6em"}
```{r, echo=FALSE}
risk_ratio <- risk_new / risk_std
risk_ratio
```

risk ratio compares the response probabilities between the two groups

$$
\frac{\hat{p}_{\text{new}}}{\hat{p}_{\text{standard}}}
$$
\hat{p}_{new}: response probability under the new therapy
\hat{p}_{standard}: response probability under standard care

$$
\frac{0.732}{0.495}}=1.48
$$
- patients receiving the new therapy are about 1.48 times as likely to respond as those receiving standard care
- risk difference [0.236] → response is 23.6 percentage points higher
- risk ratio [1.48] → response is 48% higher relative to standard care
::: 
-----------------------------------------------------------------

### Odds of Response in Each Group
::: {style="font-size:0.6em"}
```{r}
response_odds <- response_summary |>
  mutate(odds = risk / (1 - risk))

response_odds
```



> How do the odds of response compare between the two groups?

**Interpretation**

$$
odds = \frac{p}{1-p}
$$

Where:

**p**: probability of response
**1−p**: probability of no response

New therapy [odds = 2.73]
  - odds of response are ~ 2.7
  - response is much more likely than non-response

Standard care [odds = 0.98]
  - odds are about 1 
  - response and non-response are likely to occur at similar rates


:::
-----------------------------------------------------------------

### Estimated Odds Ratio for Clinical Response
::: {style="font-size:0.6em"}
```{r, echo=FALSE}
odds_ratio <- response_odds$odds[response_odds$treatment == "New therapy"] /
  response_odds$odds[response_odds$treatment == "Standard care"]

odds_ratio
```

Goal: estimate the relative odds of response under the new therapy compared with standard care

Interpretation: odds of clinical response are approximately 2.78 times higher under the new therapy compared to standard care
::: 
-----------------------------------------------------------------



### Summary
::: {style="font-size:0.6em"}

- discussed different metrics used to estimate how outcomes differ between groups in the population

Comparative estimation follows a common structure:

1. estimate the outcome within each group
2. combine those estimates into a comparison
3. interpret the result as a population effect

Key measures:

  - mean difference → difference in averages (continuous outcomes)
  - risk difference → absolute difference in probabilities
  - risk ratio → relative difference in probabilities
  - odds ratio → relative difference on the odds scale

::: 

