---
title: "Understanding Estimates Through Uncertainty"
author: "Lesley Chapman Hannah, Ph.D., M.S."

format: pptx

execute:
  echo: true
  warning: false
  message: false
---

```{r setup, echo=FALSE}
library(tidyverse)
library(scales)

set.seed(314)

ckd_df <- tibble(
  id = 1:120,
  treatment = rep(c("Standard care", "Intervention"), each = 60),
  egfr = c(
    rnorm(60, mean = 58, sd = 11),
    rnorm(60, mean = 63, sd = 11)
  ),
  response = c(
    rbinom(60, 1, 0.35),
    rbinom(60, 1, 0.52)
  ),
  age = round(rnorm(120, mean = 62, sd = 9)),
  bmi = round(rnorm(120, mean = 30, sd = 4), 1)
)

mean_ci_df <- ckd_df |>
  summarise(
    n = n(),
    mean_egfr = mean(egfr),
    sd_egfr = sd(egfr),
    se_mean = sd_egfr / sqrt(n),
    t_crit = qt(0.975, df = n - 1),
    margin_error = t_crit * se_mean,
    lower = mean_egfr - margin_error,
    upper = mean_egfr + margin_error
  )

sample_size_demo <- tibble(
  n = c(20, 40, 80, 160, 320),
  assumed_sd = 11
) |>
  mutate(
    se_mean = assumed_sd / sqrt(n),
    t_crit = qt(0.975, df = n - 1),
    margin_error = t_crit * se_mean,
    approx_width = 2 * margin_error
  )
```



### Review: Estimation in Analysis
::: {style="font-size:0.65em"}

estimation is the process of evaluating how that number behaves across repeated sampling

the following statistical estimation concepts were established:


- **population quantity [parameter]** : defines the scientific target

  - example: mean eGFR, disease prevalence

- **sample-based estimator**: provides a way to approximate that quantity

  - example: sample mean, sample proportion

- **point estimate** : summarizes observed data into a single value

Evaluation of estimators using two key properties:

  - bias [accuracy]
    
    - whether the estimator is centered on the true parameter         across repeated samples

  - variance [precision]

    - how much the estimate varies from one sample to another

::: 
---

### Extending Estimation

::: {style="font-size:0.6em"}

- previously we discussed that estimation is a task that involves using sample data to learn about an unknown population quantity
- point estimate is single  realization from a repeated-sampling process 
- estimators must be judged by their behavior across samples, including bias and variability

Current lecture addresses the following idea:

> point estimate provides a single value, but scientific conclusions require an assessment of how much uncertainty surrounds that value

:::

------------------------------------------------------------------------

### Extending Estimation

::: {style="font-size:0.6em"}


Lecture concepts:

- Interval estimates extend point estimates by accounting for sampling variability, providing a range of values that are consistent with the data rather than a single value alone

- Every estimate is subject to sampling variability, meaning that a different sample from the same population would produce a different value

- Interval estimates provide a range that reflects this variability, indicating how stable the estimate is and how close it is likely to be to the true population quantity
:::

--------


### Review: Point Estimates

::: {style="font-size:0.66em"}
**point estimate** : metric that uses sample data to produce one numerical value for a population parameter

Examples include:

-   sample mean as an estimate of a population mean
-   sample proportion as an estimate of a population proportion
-   differences in mean - observed treatment difference as an estimate of a population effect

point estimates do not show how much the estimate could vary from sample to sample

:::

------------------------------------------------------------------------

### Point Estimate Limitations

::: {style="font-size:0.67em"}

Background: suppose a study reports that the sample mean eGFR is 60.8 mL/min/1.73 m²

Unaddressed questions:

  - How much would this estimate change if the study were repeated?
  - Is the mean estimated precisely, or only roughly?
  - Are nearby values such as 59, 61, or 64 still plausible for the population mean?

- point estimates alone include no information about variability

- two studies can report the same estimate while differing in precision

- interval estimates provide a range that reflects variability and makes those differences visible
:::

------------------------------------------------------------------------


### Interval Estimates

::: {style="font-size:0.66em"}


**interval estimate** : 

  - range of values used to estimate a population parameter 

  - extends the point estimate by giving a range of plausible values for the population parameter

Point Estimate versus Interval Estimate:

-   point estimate: best single-number summary from the sample

  - Answers: what was observed in this sample?
  
-   interval estimate: range of values compatible with the sample and its uncertainty**
  
  - Answers: what does the sample suggests about the population, allowing for sampling variability?
  
:::

-------

### Defining Interval Estimates

::: {style="font-size:0.67em"}

- **interval estimate** : sample-based range of values used to estimate an unknown population parameter

- range of values computed from sample data that estimates an unknown population quantity while incorporating sampling uncertainty

Properties of interval estimates:

-   constructed from the observed sample
-   targets a population quantity, not just the sample itself
-   reflects sampling variability by indicating that the estimate could differ across repeated samples



:::

------------------------------------------------------------------------

### Defining Confidence Interval

::: {style="font-size:0.67em"}

- interval estimates involve generating a range of numbers within which the parameter is expected to lie so that within a chosen degree of confidence the true unknown parameter is capture inside the interval

- point estimates will differ from the population parameter due to the sampling error and it is difficult to approximate how close the point estimate is to the actual parameter

- **confidence interval** : 

  - interval estimate constructed from sample data along with a stated confidence level
  
  - value generated such that over repeated samples, a specified proportion of such intervals will contain the true population parameter

Example:

- 95% confidence interval (CI) - under repeated sampling the 95% CI captures the true population parameter about 95% of the time


:::

------------------------------------------------------------------------

### Confidence Intervals and Interval Estimation

::: {style="font-size:0.67em"}

`confidence interval` : frequentist method for converting a point estimate and its sampling variability into an interval estimate for a population parameter

confidence intervals within the  sequence of estimation:

1. **identify the population quantity of interest**

 Example: mean, proportion, or difference between groups

2. **compute a point estimate from the sample**

  a single value summarizing the observed data

3. **quantify sampling variability**
  
  assess how much the estimate would vary across repeated samples [standard error]


:::

------------------------------------------------------------------------

### Confidence Intervals and Interval Estimation

::: {style="font-size:0.67em"}

`confidence interval` : frequentist method for converting a point estimate and its sampling variability into an interval estimate for a population parameter

confidence intervals within the  sequence of estimation:


4. **construct an interval estimate**

  represent that variability as a range of plausible values
  
5. **interpret the interval estimate**

  use the confidence interval to identify which values of the       population parameter are consistent with the observed data and    to assess the precision of the estimate

:::

------------------------------------------------------------------------




### Confidence Interval Versus Standard Deviation

::: {style="font-size:0.67em"}

- **standard deviation** : variability among observed individual values in the sample

- **confidence interval** : range for the population parameter based on the sample estimate and its uncertainty

Examples:

-   **standard deviation**: spread of patient-level values

  e.x.: eGFR values vary greatly from patient to patient, that affects the standard deviation
  
-   **confidence interval**: plausible range for the population parameter

  e.x.: how precisely the mean eGFR has been estimated
  
:::

------------------------------------------------------------------------



### Simulated Biomedical Dataset

::: {style="font-size:0.55em"}
```{r}
ckd_df |>
  slice_head(n = 6)
```

- dataset represents a simulated cohort of patients with chronic kidney disease (CKD) $\rightarrow$ condition characterized by a gradual decline in kidney function over time

- one row per patient

- `egfr` is a continuous measure of kidney function

- `response` is a binary indicator of clinical response

- `treatment` identifies the patient group



:::

------------------------------------------------------------------------

### Constructing a 95% Confidence Interval 

::: {style="font-size:0.55em"}
```{r}
library(tidyverse)

summary_df <- ckd_df |>
  summarise(
    n = n(),
    mean_egfr = mean(egfr),
    sd_egfr = sd(egfr),
    se_egfr = sd_egfr / sqrt(n),
    lower = mean_egfr - 1.96 * se_egfr,
    upper = mean_egfr + 1.96 * se_egfr
  )

summary_df
```
**Interpretation**

construct a 95% confidence interval for the population mean eGFR:

-   `n`[120]: sample size
-   `mean_egfr`[59.79] : point estimate of population mean
-   `sd_egfr`[11.31] : sample standard deviation $\rightarrow$ patient-to-patient spread
-   `se_mean`[1.03] : standard error of the mean $\rightarrow$ estimate-to-estimate variability

-   `lower` and `upper`[57.77, 61.81]: confidence interval endpoints - range of plausible values for the population mean eGFR

:::

------------------------------------------------------------------------



### Visualizing the Mean and Its Confidence Interval

::: {style="font-size:0.55em"}
```{r, echo=FALSE}
#| fig-width: 4
#| fig-height: 3
#| 
plot <- ggplot(summary_df, aes(x = 1, y = mean_egfr)) +
  geom_point(size = 1.5) +   # smaller dot
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.1) +
  labs(
    title = "Mean eGFR with 95% Confidence Interval",
    x = "",
    y = "eGFR"
  ) +
  coord_cartesian(ylim = c(0, NA), xlim = c(0, 2)) +
  theme_bw()

plot
```

**Interpretation**

- plot shows the estimate of the mean and the uncertainty attached to that estimate
- point : sample mean
- vertical bar : confidence interval

- interval is narrow : mean has been estimated with relatively high precision

- Note: if the interval is wide: $\uparrow$ sampling uncertainty about the population mean

:::

------------------------------------------------------------------------

### Sample Size, Standard Error, and Confidence Intervals

::: {style="font-size:0.6em"}
How Sample Size Affects Precision and Uncertainty

Standard error is defined as:

$$
SE(\bar{x}) = \frac{s}{\sqrt{n}}
$$

- as n increases → *SE* decreases
- as standard error decreases → estimates vary less across repeated samples


:::

------------------------------------------------------------------------

### Sample Size, Standard Error, and Confidence Intervals

::: {style="font-size:0.6em"}


Confidence intervals are constructed using the standard error:

$$
Estimate±(\text{critical value})×SE
$$

- smaller SE → narrower confidence intervals
- larger SE → wider confidence intervals
- larger samples → more stable estimates and tighter intervals
- smaller samples → more variability and wider intervals
- in research, larger studies produce more precise estimates of population quantities

:::

------------------------------------------------------------------------

### Confidence Intervals and Sample Size

::: {style="font-size:0.6em"}
```{r, echo=FALSE}
set.seed(29)

sample_sizes <- c(30, 100, 300, 1000)

ci_df <- map_df(sample_sizes, function(n) {
  sample_data <- ckd_df |> sample_n(n, replace = TRUE)
  
  tibble(
    n = n,
    mean = mean(sample_data$egfr),
    se = sd(sample_data$egfr) / sqrt(n),
    lower = mean - 1.96 * se,
    upper = mean + 1.96 * se
  )
})

ci_df
```

Hypothesis as `n` increases:

-   `se_mean` decreases [mean eGFR]
-   `margin_error` decreases
-   `approx_width` of the interval decreases

:::

------------------------------------------------------------------------

### Plot: Confidence Intervals vs Sample Size

::: {style="font-size:0.64em"}
```{r, echo=FALSE}

#| fig-width: 4
#| fig-height: 3


plot <- ggplot(ci_df, aes(x = n, y = mean)) +
  geom_point() +
  geom_errorbar(aes(ymin = lower, ymax = upper)) +
  labs(
    title = "Confidence Interval Width vs Sample Size",
    x = "Sample Size",
    y = "Mean eGFR"
  ) +
  theme_bw()

plot
```

**Interpretation**

- plot shows how the estimate of mean eGFR and its associated 95% confidence interval change as sample size increases
:::

------------------------------------------------------------------------

### Plot: Confidence Intervals vs Sample Size

::: {style="font-size:0.64em"}


**Plot Interpretation**


- point is a sample mean eGFR
- vertical line is the 95% confidence interval around that estimate
- estimates cluster around a similar value (~60 eGFR)
- as sample size increases $\rightarrow$ confidence intervals become narrower
- center of the estimates remains stable across sample sizes
:::

------




### Putting Together Key Concepts

::: {style="font-size:0.68em"}

1.  A **point estimate** gives one numerical estimate of a population quantity
2.  An **interval estimate** adds uncertainty by giving a range of plausible values
3.  A **confidence interval** is a specific kind of interval estimate with a repeated-sampling confidence interpretation
4.  Larger samples reduce the standard error $\rightarrow$ narrowing the confidence interval and improving precision

:::

------------------------------------------------------------------------

### Summary

::: {style="font-size:0.70em"}
- statistical analysis often involves estimating population quantities using sample data

across repeated samples:

- estimates vary due to **sampling variability**  
- estimators can be evaluated by:
  - **bias** (accuracy)
  - **variance** (precision)

- point estimates summarize observed data but do not reflect uncertainty.

  
:::


---

### Summary

::: {style="font-size:0.70em"}


- **standard error (SE)** quantifies estimate variability  

$$
SE = \frac{s}{\sqrt{n}}
$$

- **confidence intervals (CI)** provide a range of plausible values:

$$
\text{Estimate} \pm (\text{critical value}) \times SE
$$


- larger sample size → smaller SE → narrower intervals  
- intervals reflect precision of the estimate  
  
:::


---

